---
title: "Data 609: Final Project"  
subtitle: "Basic Portfolio Optimization"
author: "Aaron Grzasko"
date: "December 5, 2017"
output: ioslides_presentation
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, comment=NA, message=FALSE, cache=TRUE )
```
  
```{r results='hide'}
library(knitr)
library(quantmod)
library(iterpc)
library(DT)
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)
library(ggrepel)
library(lpSolve)
library(corrplot)
library(quadprog)

```
## Intro {.smaller}  

We have eight years of monthly price data for nine passively-managed funds we're considering for our retirement portfolio.  How might we use the historical data to determine the optimal investment allocation in each fund?   

```{r}
# input and print available funds 
fund_names <- c("Vanguard Large Cap Index Fund", "Vanguard Value Index Admiral Fund",
                "Vanguard Small Cap Value ETF","Vanguard REIT Index Admiral Fund",
                "iShares MSCI EAFE Value ETF","iShares MSCI EAFE Small Cap ETF",
                "Vanguard Emerging Markets Stock Index Admiral Fund", 
                "iShares Barclays Short Treasury Bond ETF", 
                "SPDR Barclays Capital Short-Term International Treasury Bond ETF")

symbols <- c("VLCAX","VVIAX", "VBR", "VGSLX", "EFV", "SCZ", "VEMAX", "SHV", "BWZ")

style <- c("large cap blend", "large cap value", "small cap value", "real estate","intl large value", "intl small cap blend","emerging mkts lg blend","short term treasury","intl short-term treasury")

df <- data.frame(ticker=symbols, style=style, fund_name=fund_names)
knitr::kable(df[,1:2], col.names = c("ticker","style"))  
```  
  
## Overview  
  
- Data Pull
- Graphical Analysis  
- Global Minimum Variance Portfolio
- Maximize Return Given Risk Appetite  
    + Monte Carlo with Bootstrapping  
    + Normally Distributed Returns  
    + Linear Programming  
    + Quadratic Programming  
  
## Data Retrieval  
  
- Used the **quantmod** package to access Yahoo! Finance price data.  
- `getSymbols()` retrieves historical prices.  
- `adjustOHLC()` adjusts raw price data for dividends and splits.  
- `monthlyReturn()` converts price data to monthly, percentage returns.  

## Price History  
  
```{r results='hide'}
# fund symbols
symbols <- c("VLCAX","VVIAX", "VBR", "VGSLX", "EFV", "SCZ", "VEMAX", "SHV", "BWZ")

# set historical start and end dates: pull 8 eight years of data, 
beg = '2009-10-30' # last trading day in Oct 2009
end = '2017-10-31'  

# save time series daily for each fund as separate time series object
getSymbols(symbols, src = "yahoo", from = beg, to = end)

# calculate monthly return data
ret_matrix <- matrix(NA,nrow = 96, ncol=0)
for (i in symbols) {
    temp_col <- as.numeric(monthlyReturn(adjustOHLC(eval(parse(text=i)),use.Adjusted=TRUE))[2:97])
    ret_matrix <- cbind(ret_matrix,i=temp_col)
}
colnames(ret_matrix) <- symbols
```  
  
```{r}
# get price history of each fund and store in df
price_df <- data.frame(date=as.Date(matrix(NA, nrow = 2014)))

for (i in symbols) {
    temp_col <- as.numeric(eval(parse(text=i))[,6])
    price_df[, i] <- temp_col
}

price_df[,'date'] <- as.Date(index(temp_col))

price_df <- price_df %>% gather(fund, price,2:10) # long format

# plot price history
ggplot(price_df, aes(date,price, col=fund)) + geom_line() + theme_bw() + 
    labs(title = 'Adjusted Daily Prices', y = "price per share ($)", subtitle = "Nov 2009 - Oct 2017")

```  

## Monthly Returns  
```{r}
# monthly returns stored to df, long format
ret_df <- data.frame(date = seq(as.Date("2009/11/1"), as.Date("2017/10/1"), "months"), ret_matrix)
ret_df <- ret_df %>% gather(fund, return,2:10) 

# plot
ggplot(ret_df, aes(date, return*100)) + geom_line(aes(col=fund)) + 
  facet_wrap(~fund, ncol=3) + labs(y = "return (%)") + 
  theme_bw() + theme(panel.spacing = unit(1, "lines")) + 
  labs(title="Adjusted Monthly Returns")

```  

## Risk vs. Reward  

```{r}
# calculate means and standard deviation, and covariance of each fund
mean_vec <- apply(ret_matrix, 2, mean)
sd_vec <- apply(ret_matrix,2, sd)
cov_matrix <- cov(ret_matrix)

# risk reward df
rr_df <- data.frame(fund=names(mean_vec),return=mean_vec*100,risk=sd_vec*100, row.names=NULL )

# plot
ggplot(rr_df, aes(risk,return, col=fund)) + geom_point() + theme_bw() + 
  geom_text_repel(mapping = aes(risk, return), data = rr_df, label = rr_df$fund) + 
  labs(title = "Mean Monthly Returns vs. Standard Deviation", 
       subtitle = "from Nov 2009 - Oct 2017", x="std deviation (%)", y= 'mean return (%)')

```  
  
## Correlations{.smaller}  
  
```{r}
corrplot(cor(ret_matrix, use="complete.obs", method='pearson'), type = "upper", order = "hclust")
        

```

## Global Minimum Variance Portfolio {.smaller}  
  
This is a problem with a non-linear objective and one equality constraint.  We can solve analytically using method of Lagrange multipliers.  
  
Define:  
$\sigma_{x,P}^2=$ portfolio variance  
$\mathbf{x}=$ vector of portfolio weights  
$\Sigma=$ covariance matrix  
$\lambda=$ Lagrange multipler  
  
Objective:  

$$\min_{\mathbf{x}}\sigma_{x,p}^2=\mathbf{x^T}\Sigma\mathbf{x~~\textrm{s.t.}~~\mathbf{x^T1}=1}$$   
  
Set up the Lagrangian function as follows:  
  
$$L(\mathbf{x},\lambda) = \mathbf{x^T}\Sigma\mathbf{x} + \lambda(\mathbf{x^T1}-1)$$ 
  
## GMVP/cont {.smaller}  
  
Take the partial derivatives with respect to each asset, $x_i$ and set equal to zero:  
$$0=\frac{\partial L}{\partial x_i}=2x_i\sigma_i^2 + 2\sum_{i \neq j}\sigma_{ij} + \lambda$$
Also take the partial derivative with respect to $\lambda$ and set equal to zero:  
  
$$0 = \frac{\partial L}{\partial \lambda}=\mathbf{x^T1}-1$$  
Solve 10 equations with 10 unknowns:  
 
```{r}
# matrix A
A <- rbind(cbind(2 * cov_matrix, rep(1,nrow(cov_matrix))),c(rep(1, ncol(cov_matrix)),0))

# vector b
b <- c(rep(0, nrow(cov_matrix)),1)

# solve z
z <-solve(A) %*% b

# print porfolio weights and lambda parameter
z_df <- data.frame(matrix(sprintf("%.2f%%", z*100),1,10))
names(z_df) <- c(symbols, "lambda")
print.data.frame(z_df, row.names = FALSE )  

```  
  
Solution implies shorting several assets--not realistic for retirement account.  
  
## Monte Carlo with Bootstrapping{.smaller}  
  
- Simulation using sampling with replacement.  
    + Can't simulate each asset's returns independently because of strong pairwise correlations.
    + Solution: randomly sample from historical sets of monthly returns for all assets. 
- Obective: maximize return given upper bound on standard deviaton.  
- Nonlinear objective and nonlinear constraints.
- Good brute force technique that requires minimal theory.  
- Enumerate over universe of possible weights.  Keep universe small using simplying constraints.  
    + Assume no asset gets more than 40% allocation.  
    + assume discrete asset allocations in 10% increments(i.e. 0%, 10%, etc.).  
      
## MC / cont{.smaller}  
  
$\mathbf{\mu=}$ mean vector   
$\sigma_{p,x}=$ portfolio standard deviation, scalar  
$x_i=$ portfolio weight for asset $i$  
$S=$ constant, max allowed standard deviation  

Objective:  
$$\max_{\mathbf{x}}\mathbf{x^T\mu}$$  
subject to  
$$\sum_{i=1}^Nx_i=1$$
$$0\le x_i\le0.4~\textrm{for all }i$$
$$100x_i\bmod10=0~\textrm{for all }i$$
$$\sigma_{p,x} \le S $$ 
      
## MC Example Solution{.smaller}  

```{r}
# helper function: simulated returns for each asset using MC w/ bootstrapping
bs_sim <- function(num_sims, return_matrix) {
    
    return(return_matrix[sample(nrow(return_matrix),num_sims, replace=TRUE),])
}


# helper function:  enumerate over all possible portfolio weights
port_wts <- function(wt_vec, num_assets) {
  I <- iterpc(table(wt_vec), num_assets,replace=TRUE, ordered=TRUE)
  perms <- getall(I)
  row_sums <- apply(perms,1,sum)
  return(perms[row_sums == 1,]) # only return permutations where weights add to 1
}


# function to find max return given upper bound on sd
bs_max_ret <- function(sim_matrix, wt_matrix, max_sd) {
  sim_wt <- sim_matrix %*% t(wt_matrix)
  bs_mean_vec <- apply(sim_wt,2,mean)
   bs_sd_vec <- apply(sim_wt,2, sd)  
  max_ret <- max(bs_mean_vec[bs_sd_vec <= max_sd])
  max_pos <- which(bs_mean_vec == max_ret & bs_sd_vec <= max_sd)
  wts <- as.numeric(wt_matrix[max_pos,])
  names(wts) <- symbols
  list(max_ret = max_ret, sd = bs_sd_vec[max_pos], weights=wts) 
}
   
# bootstrap efficient frontier helper function
eff_front_bs <- function(start, stop, by=0.01, sim=bs_sim_matrix, wt=wt_matrix) {
    df <- data.frame(risk=numeric(),return=numeric())
    for (i in seq(start,stop, by)) {
        results <- bs_max_ret(sim, wt, i)
        df <- rbind(df, data.frame(risk=results$sd*100,return=results$max_ret*100))
        
    }
    df
}

bs_sim_matrix <- bs_sim(1000,ret_matrix) # simulate individual asset returns
wt_matrix <- port_wts(c(0,0.1,0.2,0.3,0.4),9) # universe of potentail portfolio weights
```  
Maximize return assuming s.d. $\le$ 1.5%

```{r}
bs_max_ret(bs_sim_matrix, wt_matrix, 0.015) 
```  
  
  
## MC Efficient Frontier  

```{r}
# calculate return and sd for multiple worst case sd scenarious
df <- eff_front_bs(0.014,0.04,0.002)

#geom_smooth(col='green',se="false")
# plot
ggplot(df, aes(x=risk,y=return)) + geom_line(col="green") + 
    geom_point(data = rr_df) + 
    geom_text_repel(mapping = aes(risk, return), data = rr_df, label = rr_df$fund, col= "blue" ) + theme_bw() + 
    labs(title = "Efficient frontier", subtitle = "Using Bootstrapped Returns",
         x="std deviation (%)", y = "mean return (%)") 
```  
  
## Normally Distributed Returns{.smaller}  
  
New terms:  
$R_i=$ return for asset $i$  
$\mathbf{R}=$ vector of returns, all assets  
$\mu_{p,x}=$ mean portfolio return  
  
Assume:  
$$R_i \sim  iid~ N(\mu_i,\sigma_i^2)$$
$$cov(R_i,R_j) = \sigma_{ij}$$  
Then the return of the portfolio is  

$$\mu_{p,x}=E[\mathbf{x'R}]=\mathbf{x^T}E[\mathbf{R}]=\mathbf{x^T \mu}$$  
The variance of the portfolio is:  
  
$$\sigma^2_{p,x}=\mathbf{x^T}\Sigma\mathbf{x}$$  
  
## Normally Distributed Returns / cont{.smaller}  
  
- Assume individual assets are normally distributed random variables.  Use historical historical means, variances as and covariances as parameter inputs.
- No need for simulation, just calculated portfolio mean and variance given a set of weights.  
- Objective: maximize return given upper bound on standard deviation.  
- Same constraints used in MC problem.  
- Enumerate over universe of possible weights, using simplifying constraints.  
  
## Normal Returns Example Solution{.smaller}  
 
Maximize return assuming s.d. $\le$ 1.5%  

```{r}
norm_max_ret <- function(ret_matrix, wt_matrix, max_sd){
  
  # determine means and covariance matrix for assets 
  mean_vec <- apply(ret_matrix, 2, mean)
  cov_matrix <- cov(ret_matrix)

  p_mean_vec <- wt_matrix %*% mean_vec
  p_sd_vec <- numeric()
  for (i in 1: nrow(wt_matrix)){
      temp_sd <- sqrt(t(matrix(wt_matrix[i,])) %*% cov_matrix %*% wt_matrix[i,])
      p_sd_vec <- c(p_sd_vec, temp_sd)
      
  }
  
  max_ret <- max(p_mean_vec[p_sd_vec <= max_sd])
  max_pos <- which(p_mean_vec == max_ret & p_sd_vec <= max_sd)
  wts <- as.numeric(wt_matrix[max_pos,])
  names(wts) <- symbols
  list(max_ret = max_ret, sd = p_sd_vec[max_pos], weights=wts)
  
}
  
norm_max_ret(ret_matrix,wt_matrix,0.015)

```
  
## Normal Returns Efficient Frontier  
  
```{r}
# normal dist efficiient fronter helper function
eff_front_n <- function(start, stop, by=0.01,ret = ret_matrix, wt=wt_matrix) {
    df <- data.frame(risk=numeric(),return=numeric())
    for (i in seq(start,stop, by)) {
        results <- norm_max_ret(ret, wt, i)
        df <- rbind(df, data.frame(risk=results$sd*100,return=results$max_ret*100))
        
    }
    df
}


# calculate return and sd for multiple worst case sd scenarious
df <- eff_front_n(0.014,0.04,0.002)

# plot
ggplot(df, aes(x=risk,y=return)) + geom_line(col='green') + 
    geom_point(data = rr_df) + 
    geom_text_repel(mapping = aes(risk, return), data = rr_df, label = rr_df$fund, col= "blue" ) + theme_bw() + 
    labs(title = "Efficient frontier", subtitle = "Assuming Normally Distributed Returns",
         x="std deviation (%)", y = "mean return (%)") 

```  

## Linear Programming{.smaller}  

- Cannot use variance or standard deviation as constraint because non-linear.  
- Compromise: start with mean absolute deviation as constraint:  
$$\frac{1}{T}\sum_{t=1}^T\Big|\sum_ix_i\big(R_i(t)-\mu_i\big)\Big| \le M\textrm{ where M constant}$$  
  
- Break out constraint above to satisfy linearity requirement:  
$$\sum_ix_i\big(R_i(t)- \mathbf{\mu_i}\big) \le y_t \textrm{ for all }t$$  
$$\sum_ix_i\big(R_i(t)- \mathbf{\mu_i}\big) \ge -y_t \textrm{ for all }t$$ 

$$\frac{1}{T}\sum_{t=1}^{T}y_t \le M\textrm{ }$$  

## Linear Programming / cont{.smaller}  
  
$$\max \mu_{p,x}= \sum_i x_i \mu_i$$  
subject to  
$$-y_t \le \sum_ix_i\big(R_i(t)- \mathbf{\mu_i}\big) \le y_t \textrm{ for all }t$$  
$$\frac{1}{T}\sum_{t=1}^{T}y_t \le M\textrm{ }$$  
$$\sum_{i}x_i = 1$$  
$$0 \le x_i \le 0.4,~\textrm{for all }i,~~y_t \ge0,~\textrm{for all }t$$  

## Linear Programming Example{.smaller} 
  
- Maximize return assuming mean absolute deviation $\le$ 1.5%.  
- Use `lp()` from **lpSolve** package.
  
```{r}
# function to maximize return given upper bound on mean absolute deviation
opt_wt <- function(risk_aversion, return_matrix, max_wt) {
    
    
    num_pd <- nrow(return_matrix) # number of periods
    num_assets <- ncol(return_matrix) # number of assets under consideration
    
    # mean return vector for all aseets
    mean_vec <- matrix(apply(return_matrix,MARGIN=2,mean))
    
    # objective function:  maximuize return
    obj <- c(mean_vec, rep(0,num_pd))
    
    # initialize matrix, left hand side for constraints
    A <- matrix(,num_pd*2,num_pd + num_assets)
   
    #  left side of constraint: - y_t + sum [ x_j * (R_j(t) - mean(x_j))] <= 0; perform for all t
    for (i in 1:num_pd) { 
        A[i,] <- c(return_matrix[i,] - t(matrix(mean_vec)), rep(0,num_pd))
        A[i, i + num_assets] <- -1
    }
    
    # left side of constraint: - y_t - sum [ x_j * (R_j(t) - mean(x_j))] <= 0; perform for all t
    for (i in (num_pd+1):(2*num_pd)) {
        A[i,] <- c(-(return_matrix[i-num_pd,] - t(matrix(mean_vec))), rep(0,num_pd))
        A[i, i - num_pd + num_assets] <- -1
    }
    
    # left hand side constraint: all weights less than max_weight
    for (i in 1:num_assets) {
        temp_row <- rep(0, num_assets + num_pd)
        temp_row[i] <- 1
        A <- rbind(A, temp_row)
    }
    
    # left side of constraint: avg y_t value less than risk aversion parameter 
    A <- rbind(A,c(rep(0,num_assets), rep(1, num_pd) / num_pd) )

    
    # left side constraint: asset weights sum to 1
    A <- rbind(A, c(rep(1,num_assets), rep(0, num_pd)))
    
    
    # constraint, left hand side:  all weights greater than 0  
    for (i in 1:num_assets) {
        temp_row <- rep(0, num_assets + num_pd)
        temp_row[i] <- 1
        A <- rbind(A, temp_row)
    }

    # constraint, left hand side:  all y_t values are greater than 0  
    for (i in 1:num_pd) {
        temp_row <- rep(0, num_assets + num_pd)
        temp_row[i+num_assets] <- 1
        A <- rbind(A, temp_row)
    }
    
    
    # constraint operators
    const_op <- c(rep("<=",num_pd*2 + num_assets + 1),"=",rep(">=",num_assets + num_pd))

    
    # right side of constraint operator
    b <- c(rep(0, num_pd*2), rep(max_wt, num_assets),risk_aversion, 1, rep(0, num_assets + num_pd))
    
    
    # solve lp problem 
    mylp <- lp("max", obj, A, const_op, b)
    
    # wt vector
    wt <- mylp$solution[1:num_assets]
    names(wt) <- symbols
    
    # compute mean abs deviation, portfolio
    p_mean <- t(matrix(wt)) %*% mean_vec
    
    mad <- numeric()
    for(i in 1: nrow(return_matrix)) {
      temp <- abs((t(matrix(return_matrix[i,])) %*% wt) - p_mean)  
      mad <- c(mad, temp)
    }
    mad <- sum(mad) / length(mad)
    
    return(list(max_ret=mylp$objval,mad=mad, weights=round(wt,2)))

}

opt_wt(0.015, ret_matrix,0.4)  
```
  
   
## Linear Programming Efficient Frontier{.smaller}  
  
```{r}

# linear programm dist efficiient fronter helper function
eff_front_lp <- function(start, stop, by=0.01,ret = ret_matrix) {
    df <- data.frame(risk=numeric(),return=numeric())
    for (i in seq(start,stop, by)) {
        results <- opt_wt(i, ret, 0.4)
        df <- rbind(df, data.frame(risk=results$mad*100,return=results$max_ret*100))
        
    }
    df
}


# calculate return and sd for multiple worst case sd scenarious
df <- eff_front_lp(0.014,0.06,0.002)


# plot
ggplot(df, aes(x=risk,y=return)) + geom_line(col='green') + 
    geom_point(data = rr_df) + 
    geom_text_repel(mapping = aes(risk, return), data = rr_df, label = rr_df$fund, col= "blue" ) + theme_bw() + 
    labs(title = "Efficient Frontier", subtitle = "Using Linear Programming",
         x="mean absolute deviation (%)", y = "mean return (%)") 

```  
  
## Quadratic Programming  
- Objective Function is quadratic; constraints are linear.  
- Maximize a combination of reward and risk by varying risk aversion parameter, $q$.  
- $q$ can vary from (0, $\infty$).  
- Use `solve.QP` from `quadprog` package.
  

## Quadratic Programming / cont{.smaller}  
  
Define:  
$q\ge 0$, constant  

  
Objective:  
  
$$\min~ \mathbf{x^T}\Sigma \mathbf{x} - q \mathbf{\mu^Tx}$$  
subject to  
$$\sum_{i}x_i = 1$$  
$$0 \le x_i \le 0.4,~\textrm{for all }i$$  
  
## Quadratic Programming Example{.smaller}  
  
Optimize with $q = 0.5$.

  
```{r}
max_wt <- 0.4


qp_func <- function(q, return_matrix, max_wt) {  
    Amat <- cbind(1, diag(ncol(return_matrix)), -diag(ncol(return_matrix)))
    bvec <- c(1, rep(0, ncol(return_matrix)),rep(-max_wt, ncol(return_matrix))) 
    Dmat <-  2*cov(return_matrix)  
    dvec <- matrix(apply(return_matrix,MARGIN=2,mean)) 
    
    
    qp_solve <- solve.QP(Dmat, dvec * q, Amat, bvec, meq=1, factorized=FALSE)  
     
    wts <- qp_solve$solution
    names(wts) <- symbols
    p_returns <- c()
    for (i in 1:nrow(return_matrix)) {
        p_returns <- c(p_returns, t(matrix(wts)) %*% return_matrix[i,])
    }
    
    std_dev <- sd(p_returns)  
    
    mean_ret <- mean(p_returns) 
    list(max_ret=mean_ret, sd=std_dev, weights=round(wts,1))
    

}


qp_func(0.5, ret_matrix, 0.4)



```  
  
## Quadratic Programming Efficient Frontier

```{r eval=FALSE}
```{r}

# quadratic programm efficiient fronter helper function
eff_front_qp <- function(start, stop, by=0.01,ret = ret_matrix) {
    df <- data.frame(risk=numeric(),return=numeric())
    for (i in seq(start,stop, by)) {
        results <- qp_func(i, ret, 0.4)
        df <- rbind(df, data.frame(risk=results$sd*100,return=results$max_ret*100))
        
    }
    df
}


# calculate return and sd for multiple worst case sd scenarious
df <- eff_front_qp(0.00,5,0.05)


# plot
ggplot(df, aes(x=risk,y=return)) + geom_line(col='green') + 
    geom_point(data = rr_df) + 
    geom_text_repel(mapping = aes(risk, return), data = rr_df, label = rr_df$fund, col= "blue" ) + theme_bw() + 
    labs(title = "Efficient Frontier", subtitle = "Using Quadratic Programming",
         x="std deviation (%)", y = "mean return (%)") 

```  
  
## Conclusion  
  
- Investment science is field that relies heavily on optimization theory to make rational decisions.      
  
- There are often multiple approaches to a given optimization problem.
  


  

  


  


  

  


  

  




  




